---
title: "Warm-up Mini-Report Mosquitoes"
author: "Mason Gagner"
date: "2025-09-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## ABSTRACT
## BACKGROUND

Fill in some text here that provides background info on the WNV system, the blood meal DNA extractions, PCR, sequencing, etc.

```{r viremia}

```
NOTE: Examples of data you can plot for the background info at 

# MAJOR QUESTION

Fill in here the question we want to answer. 

# METHODS

Fill in here, including overview of procedure and methods used for this project

## Fill in name of first analysis
```{r analysis1}
#put code here
```
## Fill in name of second analysis/plot
```{r plot}
#put code here
```

# DISCUSSION








```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
